The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2013-09-02 article.tex
Molecular fingerprints in CDK
Text and table listing all molecular fingerprints

% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}
\usepackage[citestyle=numeric-comp]{biblatex}                                                                     
\setlength{\bibitemsep}{0cm}                                                                                                                
\renewbibmacro{in:}{%                                                                                                                       
  \ifentrytype{article}{}{%                                                                                                                 
  \printtext{\bibstring{in}\intitlepunct}}}                                                                                                 
\addbibresource{bmc_article.bib}

@@

\author{
  Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
  \email{Egon L Willighagen - egon.willighagen@ki.se}
\and
  Miguel Rojas Cherto$^2$%
  \email{Miguel Rojas Cherto - steinbeck@ebi.ac.uk}
\and
  Ola Spjuth$^3$%
  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
\and
  You C B Here$^4$%
  \email{You C B Here - you.here@example.org}
\and
  C D K Project\correspondingauthor$^98$% // FIXED SECOND BUT LAST AUTHOR
  \email{C D K Project - cdk-devel@lists.sourceforge.net}
\and
  Rajarshi Guha$^{99}$% // FIXED ONE BUT LAST AUTHOR
  \email{Rajarshi Guha - steinbeck@ebi.ac.uk}
\and
  Christoph Steinbeck$^{100}$% // FIXED LAST AUTHOR
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
}

\address{
    \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, Karolinska Institutet, SE-17177 Stockholm, Sweden\\
    \iid(2)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
    \iid(3)Example Ltd., Glasgow, UK \\
    \iid(98)\url{http://cdk.sourceforge.net/} \\
    \iid(99)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
    \iid(100)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK 
}

@@

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:}
Cheminformatics is a well-established field with many applications in chemistry,
drug discovery, and the life sciences. Particularly in the latter field, however,
molecular properties are not well-understood yet, and the interaction between
molecules and larger systems remains a challenge to be modeled. Therefore, the
development of new data mining and pattern recognition continues to be important
and the need for access to a cheminformatics library that one can change remains.
\paragraph*{Results:}
We here report the 1.4 series of the Chemistry Development Kit and outline how
it evolved since the previous report. We demonstrate the new APIs we introduced
in this version and discuss the extensive quality control mechanism we have
adopted. New APIs have been introduced for substructure searching, rendering of
molecules, handling of molecular formulas, atom type perception, and InChI
support. At a quality control side we have introduced automated building,
extensive unit and use case testing, and adopted peer review.
\paragraph*{Conclusions:}
With this paper we have shown the continued effort to provide a free, Open Source
cheminformatics library, and show that such colleborative projects can survive for
a long period. We have taken advantage from the community support, and show
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computating software.
\end{abstract}

@@

Open Source cheminformatics state \ldots (EGON)

Adoption of the CDK ... Cinfony~\cite{OBoyle2008}, and other bindings \ldots ...

Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
AMBIT, ScaffoldHunter, OMG, ...

@@

Everyone is invited to contribute and write up the functionality they contributed;
only requirement: the code must have been incorporated into the main CDK library by the
time of submission of this paper.

@@

  The CDK has adopted a new atom typing framework, to split out the perception of atom types
  from the algorithms that use atom type properties. Previous CDK releases had perception
  integrated with algorithms, even though there was only one set of defined atom types.
  This led to alternative but contradiction perception algorithms, which made maintenance
  of these code bases impractical. For example, adding a new atom type required all alternatives
  to be updated.
  To solve this, a single perception class was added, called CDKAtomTypeMatcher. The actual
  algorithm used is based
  on heuristic rules, and alternative approach for perception exist. Possibly, in the future
  alternative algorithms may be implemented, while that would again increase the maintenance
  burden.

  The new class can perceive the atom type of single atoms, but also for a full atom container.
  The CDKAtomTypeMatcher class is written to handle various amounts of missing information,
  and take into account as much information from the input. Perception

@@

  CDK 1.4.XXXX [CHECK] introduces a new API for stereochemistry information: the interfaces
  IStereoChemistryElement was introduced with

@@

  Initiated by ...

@@

@@

@@

@@

  \subsubsection*{Peer review}

Another big changes since the CDK 1.2 series, is that the project adopted a peer-review
step in its development. New functionality is now developed independently from the main
CDK library until it is finished. When it is finished, the new code is send around to
other developers to be reviewed, very much like how research papers are peer-reviewed
by journals before publication. Independent developers review the code to check if it 
meets our community's standards.

@@

%%%%%%%%%%%%%%%%%%
\section*{Methods}
  \subsection*{Peer Review: Automated}
    \subsubsection*{PMD}
    \subsubsection*{JavaDoc Checks}
  \subsection*{Peer Review: Manual}
  \subsection*{Unit Testing}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  Text for this section \ldots

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
\printbibliography
}

@@

%% Use of \listoftables is discouraged.
%%
%\section*{Tables}
%  \subsection*{Table 1 - Sample table title}
%    Here is an example of a \emph{small} table in \LaTeX\ using  
%    \verb|\tabular{...}|. This is where the description of the table 
%    should go. \par \mbox{}
%    \par
%    \mbox{
%      \begin{tabular}{|c|c|c|}
%        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
%        A1 & B2  & C3 \\ \hline
%        A2 & ... & .. \\ \hline
%        A3 & ..  & .  \\ \hline
%      \end{tabular}
%      }
%  \subsection*{Table 2 - Sample table title}
%    Large tables are attached as separate files but should
%    still be described here.

>>>>

% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amssymb}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}
\usepackage{cite}

@@

\author{
  Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
  \email{Egon L Willighagen - egon.willighagen@maastrichtuniversity.nl}
\and
  John May$^2$% // FIXED SECOND AUTHOR
  \email{John May - COMPLETE@ebi.ac.uk}
\and
  Jonathan Alversson$^3$%
  \email{Jonathan Alversson - COMPLETE}
\and
  Arvid Berg$^3$%
  \email{Arvid Berg - COMPLETE}
\and
  Nina Jeliazkova$^4$%
  \email{Nina Jeliakvova - COMPLETE}
\and
  Dazhi Jiao$^5$%
  \email{Dazhi Jiao - COMPLETE}
\and
  Syed Asad Rahman$^6$%
  \email{Syed Asad Rahman - COMPLETE}
\and
  Mark Rijnbeek$^2$%
  \email{Mark Rijnbeek - COMPLETE}
\and
  Miguel Rojas Cherto$^7$%
  \email{Miguel Rojas Cherto - COMPLETE}
\and
  Ola Spjuth$^3$%
  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
\and
  Gilleain Torrance$^5$%
  \email{Gilleain Torrance - COMPLETE}
\and
  Rajarshi Guha$^{8}$% // FIXED ONE BUT LAST AUTHOR
  \email{Rajarshi Guha - COMPLETE}
\and
  Christoph Steinbeck$^{2}$% // FIXED LAST AUTHOR
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
}

\address{
    \iid(1)Department of Bioinformatics - BiGCaT, Maastricht University, Maastricht, NL-6200 MD, The Netherlands \\
    \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, 751 24 Uppsala, Sweden \\
    \iid(4)Ideaconsult Ltd, A. Kanchev 4, Sofia 1000, Bulgaria \\
    \iid(5)Example Ltd., Glasgow, UK \\
    \iid(6)Example Ltd., Glasgow, UK \\
    \iid(7)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
    \iid(8)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA
}

@@

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:}
Cheminformatics is a well-established field with many applications in chemistry,
biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
become a widely used Open Source cheminformatics toolkit, providing various models to represent
chemical structures, of which the chemical graph is essential. However,
in the first five years of the project increased so much in size that
interdependencies between components grew unmanageable large, resulting
in unpredictable instabilities.
\paragraph*{Results:}
We here report the improvements of the CDK since the 1.2 release series made
to accomodate both the increased complexity of the library, as well as
siginificant improvements of and additions to the functionality of the
library. We here outline how the CDK evolved since the previous report and
discuss the quality control approached we have adopted to ensure
stability. Additionally, new APIs have been introduced for atom type
perception, substructure searching, rendering of molecules, and handling of
molecular formulas.
\paragraph*{Conclusions:}
With this paper we have shown the continued effort to provide a free, Open Source
cheminformatics library, and show that such colleborative projects can survive for
a long period. We have taken advantage from the community support, and show
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computating software.
\end{abstract}

@@

Open Source cheminformatics has made significants steps forward recently~\cite{OBoyle2011}.
The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011} and KNIME\cite{Beisken2013}.
Furthermore, projects with specific functionality have emerged building on the CDK, including
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
and SMSD~\cite{}. Other tools used previous versions already, and follow
newer releases of the CDK, such as Bioclipse~\cite{} and AMBIT~\cite{jeliazkova2011ambit}.

However, previous version of the CDK were written by a community with specific applications
in mind, of which structure elucidation was one. It could therefore be rather competitive
in some cheminformatics domains, such as fingerprinting~\cite{Ekins,Cannon}, in other parts
it could be inadequate, such in the capturing of stereochemistry.

Continued community development of the CDK in the last few years, however, has shown that
Open Source software is here to stay. The adoption of automatic build systems and
quality control methodologies such as unit testing, automated source code validation,
and peer review by fellow developers have greatly improved the stability of the library,
At the expense of slowing down the development, it has allowed for cleaning up interdependencies
between modules of code, and has reset the scalability of the development model,
allowing for a lot of new functionality in code APIs while maintaining the quality
of code depending on those core APIs.

This has resulted in important improved functionality, such as a separately published
InChI functionality~\cite{SpjuthXXXX} and greatly improved ring detection functionality~\cite{MayYYYY},
but also a core atom type perception module, covering a much wider set of elements than
previous CDK versions and capturing many more charged and radical atom species,
a more comprehensive fingerprinting API, many speed and stability improvements, and more.

@@

We here outline various new and improved APIs in the CDK library since the two previous
publications.

@@

  Atom type perception is a core cheminformations functionality: the atom types describe
  chemical features of atoms, such as the number of neighbors, possible formal charges,
  (approximate) hybridization, electron distribution over orbitals, etc. However, the
  CDK commonly integrated atom type perception into algorithms, resulting in multiple
  and diverging copies of similar atom type schemes. With the need to add many
  new atom types for previously uncovered element, but also charged and radical atom
  types, a new approach was needed.

  This CDK version has isolated a new atom typing framework, removing the perception of
  atom types from the algorithms, allowing the perception to be properly unit tested.
  The new code defined the atom type using an atom type ontology based on the XXXXX.

  A reference perception class, CDKAtomTypeMatcher, has been written that perceives these atom types, and
  validates the perception automatically against the properties defined by the ontology.
  This class handles a variety in types of missing information, as commonly resulting
  from various (file) formats; for example, it can handle undefined hydrogen counts
  and undefined double bond positions if hybridization information is provided instead.
  That makes the code complex, and at times slow, but so far shown sufficiently
  performant for many applications. Possibly, in the future
  alternative algorithms may be implemented, while that would again increase the maintenance
  burden.

  The new code base defines XXXX atom types for YYYY elements, including ZZZZ radical
  atom types. A full list of atom types is given in Additional File X.

  The code is tested by FFFF unit tests, and has been tested against ChEBI and PubChem. REFS

@@

  CDK 1.4.XXXX [CHECK] introduces a new API for stereochemistry information: the interfaces
  IStereoChemistryElement was introduced with ...

@@

  Initiated by Niels Out ...

@@

  \subsubsection*{SMILES parser}

@@

  \subsubsection*{Ring finding}

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of the class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~\ref{tab:fingerprints}.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}

\begin{itemize}
\item \textbf{Project Name}: The Chemistry Development Kit
\item \textbf{Project home page}: \url{http://cdk.sourceforge.net/}
\item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
\item \textbf{Programming language}: Java
\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
\item \textbf{License}: LGPL v2.1 or later
\item \textbf{Any restrictions to use by non-academics}: None additional
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Competing interests}
XX and YY have companies that sell solutions based on the CDK.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
Niels Out and Stefan Kuhn are acknowledged for their contributions to the project.

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{article}
}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - The molecular fingerprints in CDK}
    \begin{tabular}{lccrc}
                             & Bit version  & Count version & Since CDK version & References \\
  \cline{2-5}
  Fingerprinter              & $\checkmark$ &               &                   & \\
  EStateFingerprinter        & $\checkmark$ &               &                   & \\
  ExtendedFingerprinter      & $\checkmark$ &               &                   & \\
  GraphOnlyFingerprinter     & $\checkmark$ &               &                   & \\
  HybridizationFingerprinter & $\checkmark$ &               &                   & \\
  KlekotaRothFingerprinter   & $\checkmark$ &               &                   & \\
  LingoFingerprinter         & $\checkmark$ &               &                   & \\
  MACCSFingerprinter         & $\checkmark$ &               &                   & \\
  PubchemFingerprinter       & $\checkmark$ &               &                   & \\
  ShortestPathFingerprinter  & $\checkmark$ &               &                   & \\
  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  &                   & [pending] \\
  SubstructureFingerprinter  & $\checkmark$ &               &                   & \\
    \end{tabular}

%    Here is an example of a \emph{small} table in \LaTeX\ using  
%    \verb|\tabular{...}|. This is where the description of the table 
%    should go. \par \mbox{}
%    \par
%    \mbox{
%      \begin{tabular}{|c|c|c|}
%        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
%        A1 & B2  & C3 \\ \hline
%        A2 & ... & .. \\ \hline
%        A3 & ..  & .  \\ \hline
%      \end{tabular}
%      }
%  \subsection*{Table 2 - Sample table title}
%    Large tables are attached as separate files but should
%    still be described here.
